---
title: "animint2 GSOC Application"
author: "J. Chen"
format: html
toc: true
toc-location: left
toc-title: "Contents:"
linkcolor: "#043521"
mainfont: Roboto Condensed, sans-serif
monofont: Roboto Mono, monospace
reference-location: margin
cap-location: margin
knitr:
  opts_chunk: 
    code-overflow: wrap
    echo: TRUE
    cache: FALSE
    autodep: TRUE
    warning: TRUE
---

```{r setup}
#| include: FALSE

library(knitr)        # tidy file conversion
library(dplyr)        # tidy file manipulation
library(readr)        # tidy file reader
library(stringr)      # tidy string manipulation
library(tibble)       # tidy tables
library(animint2)     # animated ggplot2
library(kableExtra)   # make tables
```



## Introduction

Hello! If you're reading this, you're probably Toby Dylan Hocking, Faizan Khan, or someone else who's applying to GSOC 2023. (And if you're in the latter group, best of luck!)

Why apply to GSOC for the R Project? I figured it would be a good way to improve my R skills and start contributing to something open source. I've used and relied on a lot of R packages for ...

Why apply to work on `animint2`? I think interactive models and visualizations are important for exploratory science. Here's an example from my domain of expertise. In network psychopathology, an [interactive model of depression](https://ccl.northwestern.edu/netlogo/models/community/Vulnerability_To_Depression) led to the discovery of hysteresis in symptom networks. It continues to be an important part of network theory.^[D. Borsboom. [A network theory of mental disorders](https://onlinelibrary.wiley.com/doi/full/10.1002/wps.20375). _World Psychiatry_. 2017.] 

I'm never used `animint2` before, but I have used `ggplot2`. There's a family resemblance between the packages that should help me when contributing to the former.

Of course, my super top secret reason for doing this is so I can maybe eventually publish a paper with Toby and bump my Erdős number to a 4. But shhhhh. That's just between you and me.



## Easy Test: Personalities By Country

The easy test asks us to do an exercise from the [`animint2` Manual](https://rcdata.nau.edu/genomic-ml/animint2-manual/Ch00-preface.html). Or,

> Even better: use animint2 to visualize some data from your domain of expertise.

So let's take a look at some psychometrics.


### Data Wrangling

Many psychological datasets have strict rules regarding their use. Luckily, The Open-Source Psychometrics Project has graced us with [an abundance of personality data](https://openpsychometrics.org/_rawdata/). I've elected to use the Big Five dataset. [The Big Five is otherwise known as the five-factor model of personality.]{.aside} The dataset has data on participants' age, race, and gender (amongst other information). But for the purposes of this test, I'll limit myself to item responses by country and drop the other data:

```{r only-countries}
b5 <- read_tsv("b5.tsv",
               show_col_types = FALSE)
go_away <- c("race", "age", "engnat", "gender", "hand", "source")
b5c <- b5 |> select(-any_of(go_away))
```

How many countries are there in this dataset, anyway? And what about the number of responses per country? [`"(nu"` is equivalent to `NA`.]{.aside}

```{r country-count}
num_country <- b5c |> count(country, sort = TRUE) |>
  filter(!(country == "(nu"))
num_country
```

That's a lot of countries, and most of them have very few responses. Generalizability is affected if _n_ is too small, so I'm going to somewhat arbitrarily keep the six countries with the largest sample sizes and drop the rest. Sorry, Italy.

```{r six-countries}
six_countries <- b5c |> filter(country %in% c("US", "GB", "IN", "AU", "CA", "PH"))
```

Lastly, the individual item responses aren't especially useful to me. [Individual items are useful for estimating psychometric networks. But as far as I can tell, `animint2` doesn't support them.]{.aside} So I'm going to generate sumscores for each of the five personality factors and use that instead for the interactive visualization.

```{r sumscore-tbl-1}
# add id column
plus_id <- rowid_to_column(six_countries, "id") |> rowwise(id)

# generate table of sumscores
sumscores <- plus_id |> 
  summarize(country = country,
         extraversion = sum(c_across(3:12)), 
         neuroticism = sum(c_across(13:22)), 
         agreeableness = sum(c_across(23:31)), 
         conscientiousness = sum(c_across(33:42)), 
         openness = sum(c_across(43:52)),
         .groups = "keep")
```

And now we have our final dataset:

```{r sumscore-tbl-2}
#| column: screen-inset-right

sumscores
```


### Big Five Data Viz

Now that our data are cleaned, let's explore.

```{r code2country-tbl}
#| echo: FALSE
#| column: margin

iso2c <- read_csv("iso2country.csv",
                  show_col_types = FALSE)
iso2c_tbl <- kbl(iso2c) |> kable_paper(html_font = "Roboto Condensed",
                                       full_width = FALSE)
iso2c_tbl
```

We have five personality traits and six countries. How do personality traits vary by country?

```{r f-violinize}
# limit redundancy
violinize <- list(geom_violin(
  draw_quantiles = c(0.25, 0.5, 0.75)), 
  coord_flip(),
  labs(x = NULL),
  expand_limits(y = 100))
```

```{r oc}
#| column: margin
oc <- sumscores |> 
  ggplot(mapping = aes(x = country, 
                       y = openness))
oc + violinize
```

```{r cc}
#| column: margin
cc <- sumscores |> 
  ggplot(mapping = aes(x = country, 
                       y = conscientiousness))
cc + violinize
```

```{r ec}
#| column: margin
ec <- sumscores |> 
  ggplot(mapping = aes(x = country, 
                       y = extraversion))
ec + violinize
```

```{r ac}
#| column: margin
ac <- sumscores |> 
  ggplot(mapping = aes(x = country, 
                       y = agreeableness))
ac + violinize
```

```{r nc}
#| column: margin
nc <- sumscores |> 
  ggplot(mapping = aes(x = country, 
                       y = neuroticism))
nc + violinize
```

It's impressive how similar the scores are between countries. Due to one Indian individual with a sumscore of 0 for all factors, test results from India have the largest range. But the average scores between countries do not seriously differ.

Here's another look at at the similarities between test scores: a density plot for agreeableness. In this plot, we can see the literal overlap between the distribution of test scores. [Agreeableness refers to phenomena like the tendency for compliance and altruism.]{.aside} I'll come back to the inter-country overlap when I use `animint()`.

```{r density-plot}
#| column: margin
a_density <- sumscores |> 
  ggplot(mapping = aes(x = agreeableness,
                       color = country,
                       fill = country)) +
  geom_density(alpha = 0.1)
a_density
```

There is more variance between personality factors. Compared to the other factors, openness is unusually high and the distribution of scores for neuroticism is unusually wide. Maybe they're statistical artifacts related to the kinds of people attracted to the website. Or maybe it's something else. Personality science isn't my specialty in psychology---I can only speculate.


### Interacting with Personality

All right. Time for the most relevant part of the test: using `animint()` to showcase my data.

The factors in the five-factor model of personality are meant to be mostly independent from one another.[^2] That partly means we don't expect the factors, when plotted against one another, to have any particularly strong relationships. In contrast, if we obtain a Pearson's _r_ close to 1 or −1, we may seriously consider that the two factors are actually one.

<!-- footnote -->
[^2]: There are empirical and theoretical reasons to question this, but that's outside the scope of the test. For further reading, see:

    - D. van der Linden et al. [The general factor of personality](doi.org/10.1016/j.jrp.2010.03.003). _J. Res. Pers._ 2010.
    - D. Borsboom. [The attack of the psychometricians](https://link.springer.com/10.1007/s11336-006-1447-6). _Psychometrika._ 2006.
<!-- /footnote -->

What do our data show? Here's an interactive scatterplot---extraversion is plotted against neuroticism.^[Extraversion refers to phenomena like outgoingness, while neuroticism refers to pheomena like emotional stability.]

```{r ani-corr}
correlations <- ggplot() + 
  geom_jitter(data = sumscores, 
             mapping = aes(x = neuroticism, 
                           y = extraversion, 
                           color = country))

ani_corr <- animint(correlations)
ani_corr$first <- list(country = "US")
ani_corr
```

The United States is the only country currently plotted. Clicking on other countries plots their data and shows that aforementioned overlap between personalities by countries. Pretty neat. I've also made the data points semi-transparent so the 

The data look like paint splatches

The overlap between countries is again apparent.

If I'm lucky, maybe I'll become the change I want to see.

As an aside, it turns out that "`showSelected` does not work with `StatDensity`."

```{r}
#| eval: FALSE
#| warning: FALSE

Warning: showSelected does not work with StatDensity
```

### Sibling Rivalry

The second part of the easy test asks me to

> Show an example of an error that you see when animint2 is loaded/attached at the same time as standard ggplot2.

There's no conflict with just loading the packages:

```{r}
#| eval: FALSE

library(animint2)
library(ggplot2)
```


But R throws up an error message when you run code that uses that syntax. As a quick example, I'll load up a toy dataset and try to generate a bar chart.

```{r}
#| echo: FALSE
cat_pop <- read_csv("meow.csv",
                    show_col_types = FALSE)
```

```{r}
#| eval: FALSE
cat_pop <- read_csv("meow.csv",
                    show_col_types = FALSE)

cat_bar <- ggplot(data = cat_pop, 
                  mapping = aes(cat_name, popularity)) + 
  geom_bar(stat = "identity", color = purple)
cat_bar
```

What happens? 

```{r}
#| eval: FALSE
#| warning: FALSE
Warning: Incompatible methods ("+.gganimint", "+.gg") for "+"
```

In other words, `animint2` and `ggplot2` don't get along because their syntaxes overlap. Unloading or uninstalling one or the other gets rid of the problem and finally gives us our very important kitty bar chart.

```{r}
#| echo: FALSE
#| label: fig-cat-popularity
#| fig-cap: "Popularity of cat names on a scale of 1–3."

cat_bar <- ggplot(data = cat_pop, 
                  mapping = aes(cat_name, popularity)) + 
  geom_bar(stat = "identity")
cat_bar
```



## Appendix: Observations

`animint2` is a fork of `ggplot2`. While playing around with `animint2`, I noticed that this comes with some side effects:

- `ggplot2` has `geom_col()`, but not `animint2`.
- `ggplot2` has `geom_boxplot()`, but not `animint2`.
- An erroneous warning in `geom_violin` [has been fixed](https://github.com/tidyverse/ggplot2/pull/4654) in `ggplot2` but not in `animint2`.
- "`showSelected` does not work with `StatDensity`"---but I haven't been able to find documentation about this anywhere. (Maybe I missed it.)
- Extensions of `ggplot2`, like `patchwork`, don't work for `animint2`.

- "Error: Use of clickSelects and showSelected as aesthetics has been deprecated. Please use as parameters."
- "Warning: showSelected only works with position=identity"

[Chapter 7](https://rcdata.nau.edu/genomic-ml/animint2-manual/Ch07-limitations.html) of the `animint2` Manual is also buggy and moves up by itself.