---
title: "animint2 GSOC Application"
author: "J. Chen"
format: html
toc: true
toc-location: left
toc-title: "Contents:"
linkcolor: "#043521"
mainfont: FiraGo
monofont: Fira Code
reference-location: margin
cap-location: margin
knitr:
  opts_chunk: 
    code-overflow: wrap
    echo: TRUE
    cache: TRUE
    autodep: TRUE
---

```{r setup}
#| include: FALSE

library(knitr)        # file conversion
library(dplyr)        # tidy file manipulation
library(readr)        # tidy file reader
library(stringr)      # tidy string manipulation
library(tibble)       # tidyverse df's
library(animint2)     # animated ggplot2
library(kableExtra)   # make tables
```



## Introduction

Hello! If you're reading this, you're probably Toby Dylan Hocking, Faizan Khan, or someone else who's applying to GSOC 2023. (And if you're in latter group, best of luck!)

Why apply to GSOC for the R Project? I figured it would be a good way to improve my R skills and start contributing to something open source. I've used and relied on a lot of R packages for ...

Why apply to work on `animint2`? I think interactive models and visualizations are important for exploratory science. Here's an example from my domain of expertise. In network psychopathology, an [interactive model of depression](https://ccl.northwestern.edu/netlogo/models/community/Vulnerability_To_Depression) led to the discovery of hysteresis in symptom networks. It continues to be an important part of network theory.^[D. Borsboom. ["A network theory of mental disorders."](https://onlinelibrary.wiley.com/doi/full/10.1002/wps.20375) _World Psychiatry_. 2017.] 

I'm never used `animint2` before, but I have used `ggplot2`. There's a family resemblance between the packages that should help me when contributing to the former.

Of course, my super top secret reason for doing this is so I can maybe eventually publish a paper with Toby and bump my Erdős number to a 4. But shhhhh. That's just between you and me.



## Easy Test: Personalities By Country

The easy test asks us to do an exercise from the [`animint2` Manual](https://rcdata.nau.edu/genomic-ml/animint2-manual/Ch00-preface.html). Or,

> Even better: use animint2 to visualize some data from your domain of expertise.

So let's take a look at some psychometrics.


### Data Wrangling

Most psychological datasets have strict rules regarding their use. Luckily, The Open-Source Psychometrics Project has graced us with [an abundance of personality data](https://openpsychometrics.org/_rawdata/). I've elected to use the Big Five dataset. [Otherwise known as the five-factor model of personality.]{.aside} The dataset has data on participants' age, race, and gender (amongst other information). But for the purposes of this test, I'll limit myself to item responses by country and drop the other data:

```{r}
b5 <- read_tsv("b5.tsv")
go_away <- c("race", "age", "engnat", "gender", "hand", "source")
b5c <- b5 |> select(-any_of(go_away))
```

How many countries are there in this dataset, anyway? And what about the number of responses per country? [`"(nu"` is equivalent to `NA`.]{.aside}

```{r}
num_country <- b5c |> count(country, sort = TRUE) |>
  filter(!(country == "(nu"))
num_country
```

That's a lot of countries, and most of them have very few responses. Generalizability is affected if _n_ is too small, so I'm going to somewhat arbitrarily keep the six countries with the largest sample sizes and drop the rest. Sorry, Italy.

```{r}
six_countries <- b5c |> filter(country %in% c("US", "GB", "IN", "AU", "CA", "PH"))
```

Lastly, the individual item responses aren't especially useful to me. [I could use the individual items to estimate a psychometric network. But as far as I can tell, neither `animint2` nor `ggplot2` support that.]{.aside} So I'm going to generate sumscores for each of the five personality factors and use that instead for the interactive visualization.

```{r}
# add id column
plus_id <- rowid_to_column(six_countries, "id") |> rowwise(id)

# generate table of sumscores
sumscores <- plus_id |> 
  summarize(country = country,
         extraversion = sum(c_across(3:12)), 
         neuroticism = sum(c_across(13:22)), 
         agreeableness = sum(c_across(23:31)), 
         conscientiousness = sum(c_across(33:42)), 
         openness = sum(c_across(43:52)))
```

And now we have our final dataset:

```{r}
sumscores
```


### The Big Five Under Pressure

Now that our data are cleaned, let's explore.


The Big Five's origins lie in factor analysis. The factors are meant to be orthogonal---i.e., independent from one another. Do our data support that claim? I'll generate an interactive scatterplot for the reader to explore:

```{r}
correlations <- ggplot() + 
  geom_point(data = sumscores, 
             mapping = aes(x = neuroticism, 
                           y = openness, 
                           color = country))
animint(correlations)
```


### Sibling Rivalry

The second part of the easy test asks me to

> Show an example of an error that you see when animint2 is loaded/attached at the same time as standard ggplot2.

There's no conflict with just loading the packages:

```{r}
#| eval: FALSE

library(animint2)
library(ggplot2)
```


But R throws up an error message when you run code that uses that syntax. As a quick example, I'll load up a toy dataset and try to generate a bar chart.

```{r}
cat_pop <- read_csv("meow.csv",
                    show_col_types = FALSE)
```
```{r}
#| eval: FALSE

cat_bar <- ggplot(data = cat_pop, 
                  mapping = aes(cat_name, popularity)) + 
  geom_bar(stat = "identity", color = purple)
cat_bar
```

What happens?

```
Warning: Incompatible methods ("+.gganimint", "+.gg") for "+"
```

In other words, `animint2` and `ggplot2` don't get along because their syntaxes overlap. Unloading or uninstalling one or the other gets rid of the problem and finally gives us our very important kitty bar chart.

```{r}
#| echo: FALSE
#| label: fig-cat-popularity
#| fig-cap: "Popularity of cat names on a scale of 1–3."

cat_bar <- ggplot(data = cat_pop, 
                  mapping = aes(cat_name, popularity)) + 
  geom_bar(stat = "identity")
cat_bar
```


## Technical Appendix

Here are all the little technical bits that don't matter all that much, but which I noticed while taking the tests:

- The Big Five dataset is provided as a `.csv`, but it's actually a `.tsv` file.
- `ggplot2` has `geom_col()`, but not `animint2`.


