---
title: "animint2 GSOC Application"
author: "J. Chen"
format: html
toc: true
toc-location: left
toc-title: "Contents:"
linkcolor: "#043521"
mainfont: FiraGo
monofont: Fira Code
reference-location: margin
cap-location: margin
knitr:
  opts_chunk: 
    code-overflow: wrap
    echo: TRUE
---

```{r setup}
#| include: FALSE

library(knitr)        # file conversion
library(readr)        # tidyverse file reader
library(dplyr)        # tidyverse file manipulation
library(animint2)     # animated ggplot2
library(kableExtra)   # make tables
```

### Sibling Rivalry

The second part of the easy test asks me to

> Show an example of an error that you see when animint2 is loaded/attached at the same time as standard ggplot2.

There's no conflict with just loading the packages:

```{r}
#| eval: FALSE

library(animint2)
library(ggplot2)
```


But R throws up an error message when you run code that uses that syntax. As a quick example, I'll load up a toy dataset and try to generate a bar chart. [testing aside]{.aside}

```{r}
#| echo: FALSE
cat_pop <- read_csv("meow.csv")
cat_pop
```

```{r}
#| eval: FALSE

cat_bar <- ggplot(data = cat_pop, 
                  mapping = aes(cat_name, popularity)) + 
  geom_bar(stat = "identity", color = purple)
cat_bar
```

What happens?

```{r, eval = FALSE}
Warning: Incompatible methods ("+.gganimint", "+.gg") for "+"
```

In other words, `animint2` and `ggplot2` don't get along because their syntaxes overlap. Unloading or uninstalling one or the other gets rid of the problem and finally gives us our very important kitty bar chart.

```{r}
#| echo: FALSE
#| label: fig-cat-popularity
#| fig-cap: "Popularity of cat names on a scale of 1–3."

cat_bar <- ggplot(data = cat_pop, 
                  mapping = aes(cat_name, popularity)) + 
  geom_bar(stat = "identity")
cat_bar
```