---
title: "animint2 GSOC Examination"
author: J. Chen
format: 
  html:
    css: changes.css
mainfont: "Roboto Condensed"
monofont: "Roboto Mono"
number-sections: FALSE
toc: TRUE
toc-location: left
toc-title: "Contents:"
linkcolor: "#043521"
reference-location: margin
cap-location: margin
website:
  cookie-consent: express
knitr:
  opts_chunk: 
    code-overflow: wrap
    echo: TRUE
    cache: FALSE
    autodep: TRUE
    warning: FALSE
---

```{r setup}
#| include: FALSE

library(knitr)        # tidy file conversion
library(dplyr)        # tidy file manipulation
library(readr)        # tidy file reader
library(stringr)      # tidy string manipulation
library(tidyr)        # tidy uhhh tidying
library(tibble)       # tidy tables
library(animint2)     # animated ggplot2
library(kableExtra)   # make tables
library(testthat)     # unit testing

theme_set(theme_gray() + 
            theme(panel.background = 
                    element_rect(fill = "#f1f3f5")))
```



## Introduction

Hello! If you're reading this, you're probably Toby Dylan Hocking, Faizan Khan, or someone who's applying to GSOC 2023. (And if you're in the latter group, best of luck!)

Why am I applying to work on `animint2`? In my field of expertise---network psychometrics---an [interactive model of depression](https://ccl.northwestern.edu/netlogo/models/community/Vulnerability_To_Depression) led to the discovery of hysteresis in symptom networks. Later on, this became a crucial component of network theory.^[D. Borsboom. [A network theory of mental disorders](https://onlinelibrary.wiley.com/doi/full/10.1002/wps.20375). _World Psychiatry_. 2017.] So I've become an enthusiast for interactive models and visualizations in science.

If I'm lucky, I hope to spend my summer improving `animint2`. I'm especially interested in writing and editing documentation. But I'm comfortable doing whatever has priority. [Of course, my super top secret reason for doing this is so I can maybe eventually publish a paper with Toby and bump my Erdős number down to a 4. But shhhh. That's just between you and me.]{.aside}

[Here's my repository](https://github.com/ampurr/gsoc). It has my `.qmd` file, if you'd prefer to look at that or knit it yourself. Unless you also use the accompanying CSS file, there may be slight aesthetic differences between your rendered HTML file and this website.

Lastly, there's been an explosion of AI tools this year. I hereby certify that I haven't used them for any of my writing or code. In fact, I haven't used them at all.

All right. Onto the tests.



## Easy Test: Personalities By Country

The easy test asks us to do an exercise from the [`animint2` Manual](https://rcdata.nau.edu/genomic-ml/animint2-manual/Ch00-preface.html). Or,

> Even better: use animint2 to visualize some data from your domain of expertise.

So let's take a look at some psychometrics.


### Data Wrangling

Many psychological datasets have strict rules regarding their use. Luckily, The Open-Source Psychometrics Project has graced us with [an abundance of personality data](https://openpsychometrics.org/_rawdata/). I've elected to use the Big Five dataset. [The Big Five is otherwise known as the five-factor model of personality.]{.aside} The dataset has data on participants' age, race, and gender (among other things). But for the purposes of this test, I'll limit myself to item responses by country and drop the other data:

```{r only-countries}
b5 <- read_tsv("b5.tsv",
               show_col_types = FALSE)
go_away <- c("race", "age", "engnat", "gender", "hand", "source")
b5c <- b5 |> select(-any_of(go_away))
```

How many countries are there in this dataset, anyway? And what about the number of responses per country? [`"(nu"` is equivalent to `NA`.]{.aside}

```{r country-count}
num_country <- b5c |> count(country, sort = TRUE) |>
  filter(!(country == "(nu"))
num_country
```

That's a lot of countries, and most of them have very few responses. Generalizability is affected if _n_ is too small, so I'm going to somewhat arbitrarily keep the six countries with the largest sample sizes and drop the rest. Sorry, Italy.

```{r six-countries}
six_countries <- b5c |> filter(country %in% c("US", "GB", "IN", "AU", "CA", "PH"))
```

Lastly, the individual item responses aren't especially useful to me. [Individual items are useful for estimating psychometric networks. But as far as I can tell, `animint2` doesn't support them.]{.aside} So I'm going to generate sumscores for each of the five personality factors and use that instead for the interactive visualization.

```{r sumscore-tbl-1}
# add id column
plus_id <- rowid_to_column(six_countries, "id") |> rowwise(id)

# generate columns of sumscores
plus_sumscores <- plus_id |> 
  mutate(extraversion = sum(across(starts_with("E"))),
         neuroticism = sum(across(starts_with("N"))),
         agreeableness = sum(across(starts_with("A"))),
         conscientiousness = sum(across(starts_with("C", ignore.case = FALSE))),
         openness = sum(across(starts_with("O"))))

# remove individual items
sumscores <- plus_sumscores |> ungroup() |>
  select(id, country, extraversion:openness)
```

And now we have our final dataset:

```{r sumscore-tbl-2}
#| column: body-outset-right
sumscores
```


### Big Five Data Viz

Now that our data are cleaned, let's explore.

```{r code2country-tbl}
#| echo: FALSE
#| column: margin

iso2c <- read_csv("iso2country.csv",
                  show_col_types = FALSE)
iso2c_tbl <- kbl(iso2c) |> kable_paper(html_font = "Roboto Condensed",
                                       full_width = FALSE)
iso2c_tbl
```

We have five personality traits and six countries. How do personality traits vary by country?

```{r f-violinize}
# limit redundancy
violinize <- list(geom_violin(
  draw_quantiles = c(0.25, 0.5, 0.75)), 
  coord_flip(),
  labs(x = NULL))
```

```{r oc}
#| column: margin
oc <- sumscores |> 
  ggplot(mapping = aes(x = country, 
                       y = openness))
oc + violinize
```

```{r cc}
#| column: margin
cc <- sumscores |> 
  ggplot(mapping = aes(x = country, 
                       y = conscientiousness))
cc + violinize
```

```{r ec}
#| column: margin
ec <- sumscores |> 
  ggplot(mapping = aes(x = country, 
                       y = extraversion))
ec + violinize
```

```{r ac}
#| column: margin
ac <- sumscores |> 
  ggplot(mapping = aes(x = country, 
                       y = agreeableness))
ac + violinize
```

```{r nc}
#| column: margin
nc <- sumscores |> 
  ggplot(mapping = aes(x = country, 
                       y = neuroticism))
nc + violinize
```

There's some limited differences between personality factors. The distribution of scores for neuroticism is unusually wide, and openness seems to be slightly higher than the other personality factors. I'm not sure why. Maybe it's an artifact of the kinds of people attracted to the website. Personality science isn't my specialty in psychology---I can only speculate.

It's also impressive how similar the scores are between countries. Their median scores do not seriously differ.

Here's another look at at the similarities between test scores: a density plot for agreeableness. In this plot, we can see the literal overlap between the distribution of test scores. [Agreeableness refers to phenomena like the tendency for compliance and altruism.]{.aside} I'll come back to the inter-country overlap later.

```{r density-plot}
#| column: margin
a_density <- sumscores |> 
  ggplot(mapping = aes(x = agreeableness,
                       color = country,
                       fill = country)) +
  geom_density(alpha = 0.1)
a_density
```


### Interacting with Personality

All right. Time for the most relevant part of the test: using `animint()` to showcase my data.

The factors in the five-factor model of personality are meant to be mostly independent from one another.[^2] That partly means we don't expect the factors, when plotted against one another, to have any particularly strong relationships. In contrast, if we obtain a Pearson's _r_ close to 1 or −1, it may be the case that the two factors are actually one.

<!-- footnote -->
[^2]: There are empirical and theoretical reasons to question this, but that's outside the scope of the test. For further reading, see:

    - D. Borsboom. [The attack of the psychometricians](https://link.springer.com/10.1007/s11336-006-1447-6). _Psychometrika._ 2006.
    - D. van der Linden et al. [The general factor of personality](doi.org/10.1016/j.jrp.2010.03.003). _J. Res. Pers._ 2010.
<!-- /footnote -->

What do the data show?

```{r ani-corr}
#| message: FALSE
#| label: fig-personality-us
#| fig-cap: "The US's data seriously overlap with other countries'."

correlations <- sumscores |> 
  ggplot(mapping = aes(x = conscientiousness,
                       y = extraversion,
                       color = country))
corr <- correlations + geom_jitter(alpha = 0.25)

ani_corr <- animint(corr)
ani_corr$first <- list(country = "US")
ani_corr
```

Here, extraversion is plotted against conscientiousness. [Extraversion refers to phenomena like outgoingness, while conscientiousness refers to phenomena like diligence.]{.aside} The United States is the only country currently plotted. Clicking on other countries plots their data and shows that aforementioned overlap between countries. I've also made the data points semi-transparent so the data density is visible.

I haven't bothered to do any statistics. But at first glance, it seems like these data are consistent with the five-factor model. Outliers distort our hypothetical regression line and render $r > 0$, but if we zoom in where the data are most dense, perhaps even that fluke of a pattern will fade.

Let's see if that's true.

```{r ani-corr-zoom}
#| message: FALSE
#| label: fig-personality-ph
#| fig-cap: "There seems to be no relationship between the two factors."

corr_densed <- corr + 
  xlim(20, 40) + 
  ylim(20, 40)

ani_corr_zoom <- animint(corr_densed)
ani_corr_zoom$first <- list(country = "PH")
ani_corr_zoom
```

This time I've opted to show the Philippines, the country with the sparsest data. The data do seem to be pretty patternless. Furthermore, as you add more countries---or just the United States---the discrete nature of the data becomes visible even with jittered points. (The sumscores can only be integers.)

Of course, it's immediately obvious without jitter. In the following heatmap, we can see both the discreteness and distribution of the data. [Thanks to Toby Dylan Hocking for the suggestion to use a heatmap here!]{.aside} However, generating the heatmap is not as straightforward as generating a scatterplot. First, we need to generate all combinations between the extraversion and conscientiousness factors. Then we need to count them:

```{r heatmap-tbl}
combo_tbl <- sumscores |> 
  select(extraversion, conscientiousness) |>
  expand_grid(.name_repair = "minimal")

von_count <- combo_tbl |> 
  group_by(across(extraversion:conscientiousness)) |>
  count()
```

Finally, we can generate our heatmap:

```{r fig-personality-discrete}
#| label: fig-personality-discrete
#| fig-cap: "Discreteness and density of the Big Five data."

corr_discrete <- von_count |>
  ggplot(mapping = aes(x = conscientiousness,
                       y = extraversion,
                       fill = n)) +
  geom_tile() + coord_fixed()
corr_discrete
```

Pretty cool! Being able to explore and present this kind of research would be harder without the interactive benefits of `animint2`.


### Sibling Rivalry

The second part of the easy test asks me to

> Show an example of an error that you see when animint2 is loaded/attached at the same time as standard ggplot2.

There's no conflict with just loading the packages:

```{r}
#| eval: FALSE
library(animint2)
library(ggplot2)
```

But R throws up an error message when you run code that uses syntax common to both. As a quick example, I'll load up a toy dataset and try to generate a bar chart.

```{r cat-tbl}
#| echo: FALSE
cat_pop <- read_csv("meow.csv",
                    show_col_types = FALSE)
```

```{r}
#| eval: FALSE
cat_pop <- read_csv("meow.csv",
                    show_col_types = FALSE)

cat_bar <- ggplot(data = cat_pop, 
                  mapping = aes(cat_name, popularity)) + 
  geom_bar(stat = "identity", color = purple)
cat_bar
```

What happens? 

```{r}
#| eval: FALSE
#| warning: FALSE
Warning: Incompatible methods ("+.gganimint", "+.gg") for "+"
```

In other words, `animint2` and `ggplot2` don't get along because their syntaxes overlap. Unloading or uninstalling one or the other gets rid of the problem and finally gives us our very important kitty bar chart.

```{r fig-cat-pop}
#| echo: FALSE
#| label: fig-cat-popularity
#| fig-cap: "Popularity of cat names on a scale of 1–3."

cat_bar <- ggplot(data = cat_pop, 
                  mapping = aes(cat_name, popularity)) + 
  geom_bar(stat = "identity")
cat_bar
```



## Medium Test: The Law of Large Numbers

There are multiple medium tests. The first one requires that the test-taker

> translate an example of the animation package into an Animint.

So let's look at the law of large numbers (LLN).


### About the LLN

The law of large numbers, also called the weak law of large numbers, was named by S. D. Poisson in an article published in 1835.[^toc] He wrote, emphasis his, that

> Things of every kind are subject to a universal law that we may call _the law of large numbers_. It consists in this: if one observes a very considerable number of events of the same kind, depending on causes that vary irregularly, that is to say, without any systematic variation in one direction, then one finds that the ratios between the numbers of events are very nearly constant. 

[^toc]: Translated from French by I.&nbsp;Hacking, in [Chapter 12](https://doi.org/10.1017/CBO9780511819766.012) of _The Taming of Chance_. 1990.

    Some sources cite [Poisson's 1837 book](https://worldcat.org/title/recherches-sur-la-probabilite-des-jugements-en-matiere-criminelle-et-en-matiere-civile-precedees-des-regles-generales-du-calcul-des-probabilites/oclc/669708893) as the origin of the LLN's name. Not quite. In [his 1835 article](https://gallica.bnf.fr/ark:/12148/bpt6k29606/f478.item), he mentions "la loi des grands nombres"---the law of large numbers.

Though Poisson named it, the LLN has its origins in a posthumous publication by Jacob Bernoulli over a century earlier, and it has continued developing even up to the twentieth century.[^llnh] The exact details of the law have changed over the centuries. For our purposes, those details are largely irrelevant. (Also, I know my limits---there's no way I can explain the LLN in a manner satisfactory to a mathematician.) Instead, we'll be thinking of the LLN like this: given some population mean and some sample mean, the sample mean approaches the population mean as the sample size increases.

[^llnh]: There's some apparent disagreement on what counts as the beginning of the LLN's history:

    - M. Denker. [Tercentennial anniversary of Bernoulli’s law of large numbers](doi.org/10.1090/S0273-0979-2013-01411-3). _Bull. Am. Math. Soc_. 2013.
    - E. Seneta. [A Tricentenary history of the Law of Large Numbers](https://doi.org/10.3150/12-BEJSP12). _Bernoulli_. 2013.

Importantly, that's how the [LLN `animation` demonstration](https://yihui.org/animation/example/lln-ani/) interprets the LLN as well. I'll construct my function with this understanding of the LLN in mind.


### Function Construction

The `lln.ani` function [takes six arguments](https://github.com/yihui/animation/blob/main/R/lln.ani.R) and has useful defaults for them all. That's a good goal to have---a user should be able to run the LLN function without any arguments and produce something useful to teach with.

Let's call the `animint2` LLN function `fn_lln`. To work, `fn_lln` requires

1. The ability to read some set of data;
2. The ability to generate the population mean for that data;
3. The ability to continuously samples from the aforementioned data set and generate sample means, as well as a way to make it stop; and
4. A plot that visually indicates the population mean, graphs out the sample means as they are generated, and updates _n_ as it increases.

Furthermore, though this isn't necessary for the function to work, I want `fn_lln` to output messages that are useful for someone who's unfamiliar with R. I'll use the `assertive` package to do this instead of relying on base R's functions, which sometimes outputs arcane error messages.

The first requirement is straightforward. The dataset needs to take the form of a numeric vector. If a user tries to use something else, `fn_lln` should throw up an error. I'll call the dataset "ds" for short:

```{r fun-data}
#| eval: FALSE
validity_check <- ds |> 
  assertive::assert_is_numeric()
```

The second requirement is also straightforward.

```{r fun-mu}
#| eval: FALSE

```

The final function looks like this:

```{r fun-lln}
fn_lln <- function(distr){
  ds <- distr |> 
  assertive::assert_is_numeric()
}
```

```{r informal-test}
meow <- sample(letters, size = 10, replace = TRUE)
in_test <- fn_lln(meow)
```


### Running `fn_lln`

I also want it to have a useful default for pedagogical purposes. Six-sided dice. Coin flips are more familiar, but they're less obvious than dice, which have their numbers printed on their side.

```{r fun-data-d6}
d6 <- sample(x = 1:6, size = 10000, replace = TRUE)
```



## Hard Test: Testing `fn_lln`

Will `fn_lln` work right? 



## Appendix: Observations

Here are some things that I noticed while reading the documentation and playing around with the package. Since I'm new to `animint2`, I make no claim that my observations are novel, noteworthy, or even correct. But I think they're still useful. My guess is that many users of `animint2` are also "new" to it, in the sense that they're working scientists whose goal is to generate an interactive graphic and then to stop thinking about the package. Maybe some of my (possibly mistaken) observations overlap with theirs.


### Thoughts

I know that `animint2` is a fork of `ggplot2`, not an extension. But the family resemblance between `animint2` and `ggplot2` is sometimes confusing. For a short while, I admit that I treated the former as if it were an extension of the latter. I tried to use `geom_col()` and then `geom_boxplot()` before checking the [reference manual](https://cran.r-project.org/web/packages/animint2/) and realizing they were not part of `animint2`. I also foolishly tried to use packages that depend on `ggplot2`. Since I had `animint2` loaded instead, that obviously didn't work. [Fun fact: I recently realized that the second edition of the `ggplot2` book mentions `animint`.]{.aside}

[`ggplot2` documentation](https://ggplot2.tidyverse.org/reference/index.html) is available as a website, but [`animint2` documentation](https://cran.r-project.org/web/packages/animint2/) is only available as a PDF. [Grumpy aside: why are PDFs R's default documentation format? It's the same with $\LaTeX$, but at least there's an obvious technical reason for that.]{.aside} Of course, there's the valuable [`animint2` Manual](https://rcdata.nau.edu/genomic-ml/animint2-manual/Ch00-preface.html). But that's more like a book than a reference site---something like a counterpart to the [`ggplot2` book](https://ggplot2-book.org/). Since the `ggplot2` reference was a lot more accessible, I found myself using their reference and only going back to the `animint2` reference when necessary.

I'm not sure if I'll have enough time, and I'm not sure it's what you mean when you wrote that `animint2` will "have even more... documentation." But as part of that documentation boost, I'd like to generate a reference site for the unique features of `animint2`. I think that'd make the package more accessible to `animint2`'s users.


### Possible Bugs?

Here's a list of `animint2`'s unexpected behaviors. My mistaken observations are probably most concentrated here. It's possible that the unexpected behavior is not the fault of `animint2`, but instead the fault of another package or user error. Honestly? Probably user error.

- [Chapter 7](https://rcdata.nau.edu/genomic-ml/animint2-manual/Ch07-limitations.html) of the `animint2` Manual is buggy and scrolls up by itself. [Technically not about `animint2` itself, but it's `animint2` adjacent.]{.aside} The effect begins happening around this sentence: "Note how we used `geom_segment` instead of `geom_bar`, but their appearance is similar."
- It's difficult to lay out multiple plots in a grid, or on a line. Quarto will not knit them properly. My workaround is to have a codeblock for each graph. There's [an open issue](https://github.com/tdhock/animint/issues/150) about it for `animint`, but not for `animint2`. So I'm not sure if it's still a problem, or if I messed up somehow.
- When I tried to make an interactive density plot, I was warned that "`showSelected` does not work with `StatDensity`." Items in the legend are selectable, but selecting one removes the density curve currently on screen. The end result is that only one curve shows at a time. I haven't been able to find documentation about this anywhere.
- When I make interactive scatterplots, I'm warned that "`showSelected` only works with `position=identity`, problem: geom1_point_foobar." This is confusing, since:

    + It's already the case that `position = identity`, since it's the default for `geom_point()` and `geom_jitter()`.
    + The problem it points out is the `.tsv` file that `animint()` generates.
    + It correctly renders the scatterplot.
    + It passes `testthat::expect_no_warning()` and fails `expect_warning()`, despite the fact that it outputs a warning.

- When I make violin plots, I'm warned that the graph is "collapsing to unique 'x' values." This [has been fixed](https://github.com/tidyverse/ggplot2/pull/4654) in `ggplot2` but not in `animint2`, possibly since the fix happened after the fork. (I'm least certain about this observation.)
- When I try to place a static `animint2` graph in the margins, I can't place a caption, or else the graph doesn't render correctly. `column: margin` and `fig-cap` don't work well together.
- When I try to use `geom_bin2d()`, I'm informed that "`bin2d` is not supported in `animint`. Try using `geom_tile()` and binning the data yourself." But the `animint2` reference mentions that it does support `geom_bin2d`.
- When I use `geom_raster()`, `animint2` correctly renders it as a static graph. But it will not render the data or the legend when I use `animint2`. This [is documented](https://github.com/tdhock/animint2/issues/53).
- When I use `geom_hex`, I'm warned that there is "no data in geom1_hex_foobar."


### Notes and Solutions

- In `ggplot2`, it's possible to plot a normal distribution without any specifying any dataframes or tibbles. This is not the case for `animint2`:

```{r}
#| eval: FALSE

# works only in ggplot2
nope_normal <- ggplot() + xlim(-5,5)
nope_normal + stat_function(fun = dnorm)

# works in animint2
works_normal <- ggplot(data.frame(x = c(-5, 5)), 
                        aes(x)) + 
  stat_function(fun = dnorm)
```



## Appendix: My Mistakes

Major mistakes I made while doing these tests and then subsequently caught. This excludes mistakes re: the number of brackets, the placement of a function, or spelling.


### Easy Test

My sumscores were erroneous, but it took me an embarrassingly long time to notice and check. I first noticed that the openness sumscores were impossibly high---each sumscore has a ceiling of 50, which openness sumscores exceeded. Manually adding up the scores myself showed that extraversion had errors, too. I believe my mistake was using `c_across()` and then messing up the selection syntax, but I didn't really have the time to experiment and check.

I took a long time figuring out the details of the heatmap. The code is straightforward, but that is the result of a long process of experimentation. Most of this was because I was unfamiliar with both heatmaps and the functions necessary for combinatorics (e.g. I was unaware of `expand_grid()`). But it was a good learning experience.
